Spatial distribution of environmental DNA in a nearshore marine habitat

### 2017-01-11 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex

I felt a little more effort could have been made to facilitate the ability to replicate the analyses in this study. The methods section provided good information, but the analyses as detailed on L473 needs more clarity. An outline of the workflow steps and the input/output at each stage would be very helpful here.
\begin{response}
	%TODO email Steve about this
  In our initial draft, we provided an outline of the bioinformatics procedures in the section Methods:Sequence Data Processing (Bioinformatics), and document the procedures in painstaking detail in the Supplemental Material section Methods:Bioinformatics.
	However, we overlooked the inclusion of a direct link to the scripts used to process the raw sequence data, and have now added those.
	We certainly appreciate any encouragement to move towards more reproducible research, though we are not yet where we would like to be.
	Any specific recommendations are welcome.
	\\
\end{response}

\subsubsection*{Validity of the findings}
In response to the ``(3) Richness declines and evenness increases with distance from shore'' heading in the Discussion section, I felt that decline in richness and increase in evenness might be expected due to dilution and osmotic factors in the ocean. Perhaps some reference to tidal flushing and dilution might be pertinent here?
\begin{response}
  This is an excellent point. We have modified the relevant section to incorporate this potential mechanism.
	\\
\end{response}

I felt that under the remit of the Ecological Analyses headings throughout the study, there could have been some focus on reporting the alpha diversity and rarefaction analysis in relation to whether effective sequencing depth had been achieved for the samples in question.
\begin{response}
  We believe we adequately addressed alpha diversity by reporting richness and evenness both in a 2-dimensional and linear spatial context, with accompanying figures.
	However, we have now also added an OTU accumulation figure to the supplement and commented on it in the results.\\
\end{response}

\subsubsection*{Comments for the Author}
I thought the paper was very interesting. The study was well designed, implemented, and written. It was useful to clarify the expectations and hypotheses outlined within, however, I felt more effort could have been made to provide recommendations on how some of the limitations in eDNA sampling could be overcome.
\begin{response}
	% TODO: provide recommendations on how some of the limitations in eDNA sampling could be overcome
  Thank you for the kind words.	\\
\end{response}

### 2017-01-12 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex

\bigskip
% }

We are grateful to the editor and reviewers for their insightful feedback.
Reviewer feedback is below in regular typeface, with our responses indented and colored, including references to the line numbers in the track changes version of the manuscript.
% Additional changes to the manuscript independent of the reviews are presented at the end of this document.

%------------------------------------EDITOR-------------------------------------
\subsection*{Editor}
Both reviewers commented favourably on this manuscript and have made some useful and constructive comments which you should consider and respond to. I would also like you to respond to the comment by Bernd Haenfling on your pre-print as he is in agreement with reviewer 1 in relation to the degree of novelty this work represents (line 100 and your concluding sentence). Personally, I found your discussion of the results in the light of the advantages and limitations of this methodology interesting and highly informative. I wondered what the community analyses results would look like if you compartmentalised by e.g. life history or type (e.g. fish)?

\begin{response}
  Thank you for the positive feedback.
	We thank the commenters for pointing out recent literature that we’d missed when developing the original manuscript.
	We have now altered the language in our paper to avoid suggesting that our study is the first to do any kind of explicit spatial analysis, but (as noted below in the response to the commenter), because we have reservations about the way the referenced study was conducted, we have avoided citing it as an example.
	However, the suggestions of reviewer 1 (Wangensteen) on this matter are well taken and we have added the appropriate citations and corrected our wording.
	\\

	Like you, we were curious about the effect of the data subset you mention, and conducted that analysis using the life history data. Unfortunately, we found no meaningful patterns, and have made an attempt to clarify this in the text.
	\\
\end{response}

@@

L314-315. You can delete ``the probability that two reads chosen at random from a sample belong to different species'', since it has been already explained in the Materials and Methods section.
\begin{response}
  We made this change.\\
\end{response}

L344. ``For the first time in a marine environment, we document four key patterns''. The authors should definitely use a more modest sentence here. At least the first two patterns they claim to report ``for the first time'' have been already reported previously in several works in the marine environment. E.g.: Fonseca et al. 2014, Guardiola et al. 2015, Lallias et a. 2015, De Vargas et al. 2015, etc., although possibly never with a sampling design specifically designed to get a quantitative assessment of distance decay.
\begin{response}
  Our intent was precisely to convey the novelty of the sampling design; we failed to do that clearly.
	We removed `For the first time in a marine environment', and we have clarified the novelty of our study in the context of recently published work in the introduction.\\
\end{response}

%------------------------------------REVIEWER2----------------------------------
\subsection*{Reviewer 2 (Stephen Moss)}
\subsubsection*{Basic reporting}
There were a couple of minor grammatical errors in the manuscript. 1) The final sentence (L292-294) of the ``Sequence Data Processing (Bioinformatics)'' paragraph on page 12 in the Results section.
\begin{response}
  We corrected this.\\
\end{response}

2) The first sentence (L359) of the ``Communities far from one another tend to be less similar than those that are nearby'' paragraph on page 15 of the Discussion section.
\begin{response}
  It is unclear what the specific error is, but we acknowledge this sentence is poorly worded.
	We changed this to read `We demonstrate that distant locations have less-similar eDNA communities than proximate locations in Puget Sound, a dynamic marine environment. '\\
\end{response}

@@

The raw data is listed as available under Data Availability, though I assume this is still under embargo? The link on page 18 (L463) will need updating upon publication.
\begin{response}
  Yes; these links were a placeholder until the manuscript is accepted and have now been updated.\\
\end{response}

A GitHub DOI can be provided by Zenodo (https://zenodo.org) and will need updating upon publication.
\begin{response}
  We are aware of this and will create a GitHub version and DOI of the respository once the final version is obtained.\\
	% TODO: upon acceptance, get DOI and update
\end{response}

\subsubsection*{Experimental design}
When detailing how samples were collected (L126-128) under the Environmental Sampling header in the Methods section, I felt that information on how sampling artefacts were mediated (e.g. from L409-414) could have been provided sooner.
\begin{response}
  Perhaps we misunderstand your comment, but in the relevant part of the methods section, we state: `Each environmental sample was collected in a clean 1 liter high-density polyethylene bottle, the opening of which was covered with 500 micrometer nylon mesh to prevent entry of larger particles.'\\
\end{response}

I felt a little more effort could have been made to facilitate the ability to replicate the analyses in this study. The methods section provided good information, but the analyses as detailed on L473 needs more clarity. An outline of the workflow steps and the input/output at each stage would be very helpful here.
\begin{response}
  In our initial draft, we provided an outline of the bioinformatics procedures in the section Methods:Sequence Data Processing (Bioinformatics), and document the procedures in painstaking detail in the Supplemental Material section Methods:Bioinformatics.
	However, we overlooked the inclusion of a direct link to the scripts used to process the raw sequence data, and have now added those.
	We certainly appreciate any encouragement to move towards more reproducible research, though we are not yet where we would like to be.
	Any specific recommendations are welcome.
	\\
\end{response}

@@

\subsubsection*{Validity of the findings}
In response to the ``(3) Richness declines and evenness increases with distance from shore'' heading in the Discussion section, I felt that decline in richness and increase in evenness might be expected due to dilution and osmotic factors in the ocean. Perhaps some reference to tidal flushing and dilution might be pertinent here?
\begin{response}
  This is an excellent point. We have modified the relevant section to incorporate this potential mechanism.
	\\
\end{response}

I felt that under the remit of the Ecological Analyses headings throughout the study, there could have been some focus on reporting the alpha diversity and rarefaction analysis in relation to whether effective sequencing depth had been achieved for the samples in question.
\begin{response}
  We believe we adequately addressed alpha diversity by reporting richness and evenness both in a 2-dimensional and linear spatial context, with accompanying figures.
	However, we have now also added an OTU accumulation figure to the supplement and commented on it in the results.
	We note that our goal is not to recover an absolute measure of alpha diversity, but to make comparisons among sites for standardized sampling effort (i.e. sequencing depth).
	\\
\end{response}

\subsubsection*{Comments for the Author}
I thought the paper was very interesting. The study was well designed, implemented, and written. It was useful to clarify the expectations and hypotheses outlined within, however, I felt more effort could have been made to provide recommendations on how some of the limitations in eDNA sampling could be overcome.
\begin{response}
  Thank you for the kind words.
	We acknowledge there are many limitations to eDNA sampling; however, we believe that making more general recommendations than those we provide in the dicussion is beyond the scope of this study.
	The more general limitations and recommendations can and do receive regular, detailed attention in the many reviews published on this subject.
	Of course, if there are specific limitations you feel we have erroneously overlooked and which are within the purview of our results, we are happy to revisit this issue.
	\\
\end{response}

@@

\subsubsection*{Bernd Hänfling, Lori Lawson Handley, Dan Read and Ian Winfield}

This is well designed study and we would like to congratulate the authors to this data set and its excellent analysis. However, we feel that the authors have made an important omission in the discussion.\\

Given that the main focus of this paper is to explore the spatial distribution of community eDNA we feel that the authors should consider the results of our study on lake fish communities (Hänfling et al. 2016). In our paper we have used extensive spatial sampling of three transects across our main study lake with a total of 66 individual samples and to our knowledge there is no other data set on macrobial eDNA metabarcoding with a comparable extent of explicit sampling. Our paper confirms that the spatial distribution of individual species reflects their abundance but also shows that the pattern of distribution is not random and ecologically meaningful. Although there might be important differences between freshwater and marine habitats we feel that the results from our study are extremely relevant here and should not be omitted especially when claiming novelty of discussion point 2.\\

Hänfling B, Lawson Handley L, Read DS, Hahn C, Li J, Nichols P, Blackman RC, Oliver A, Winfield IJ (2016) Environmental DNA metabarcoding of lake fish communities reflects long-term data from established survey methods. Molecular Ecology 25, 3101-3119.
\begin{response}
	% http://onlinelibrary.wiley.com/doi/10.1111/mec.13660/full
  Thank you for your comment and for drawing our attention to your paper.
	As you note therein, the rate of publication in this field can make it difficult to keep up.
	% Our omission was not intended as a slight, merely an oversight. % given the rapid pace of publication in this field.
	We commend your extensive environmental sampling; however, upon reading your paper, we have reservations about your interpretations of the results given the molecular protocols.
  \\

	We note that you conducted replicate PCRs of each environmental sample, and pooled them prior to sequencing.
	While this practice is not uncommon in the field, it precludes the ability to test for any effects of PCR on the communities.
	We believe this is particularly problematic in the case of your study because, as far as we understand your methods, you used indexed (`tagged') primers in a single PCR reaction.
	We have shown previously that this can bias datasets in such a way that completely obscures any underlying pattern (see: http://dx.doi.org/10.1371/journal.pone.0148698).
	As we report in that study, indexed primers can bias relative read abundance by as much as 70\%.
	That is, in one environmental sample, a species may comprise 70\% of the reads in PCR replicates amplified with one indexed primer set, while being completely absent from replicates amplified by another indexed primer set.
	This would most certainly result in a significant difference among most any comparison, such as the Chi-squared test you conducted to compare lake basins.
	% If we have misunderstood your methodology, or if you have results of replication tests on the indexed primers in community samples, we would be happy to consider further the results of your study.
  \\

	Regardless, we have neutralized our statements on the novelty of our work on the basis of the points raised by your comment and those of the reviewers.

### 2017-01-12 Manuscript/v1.1_PeerJ/carkeek_grid.tex
add rarefaction
respond to reviews

%-----------------------------------NOTES---------------------------------------
% (TODO SUBMIT DATA AND METADATA; ADD URL)
% (TODO INSERT DOI FROM FINAL COMMIT BEFORE PUBLICATION)

@@

%-----------------------------------RESULTS-------------------------------------
\section*{Results}
\subsection*{Sequence Data Processing (Bioinformatics)}
Preliminary sequence analysis strongly suggested that the observed variation among environmental samples reflects true variation in the environment, rather than variability due to lab protocols, for the following reasons (note that all value ranges are reported as mean plus and minus one standard deviation). First, all libraries passed the FastQC per-base sequence quality filter, generating a total of 371,576,190 reads passing filter generated in each direction. Second, samples in this study were represented by an adequate number of reads (333,537.9 $\pm$ 112,200.5), with no individual sample receiving fewer than 130,402 reads. Third, there was a very low frequency of cross-contamination from other libraries into those reported here (5e-05$\pm$8e-05; max proportion 0.00034). Fourth, after scaling all samples to the same sequencing depth, OTUs with abundance greater than 178 reads (0.14\% of a sample's reads) experienced no turnover among PCR replicates within a sample. Fifth, sequence abundances among PCR replicates within water samples were remarkably consistent. A single sample had low similarity among PCR replicates (0.659) after removing this outlier, the lowest mean similarity among replicates within a sample was 0.966. Overall similarities among PCR replicates within a sample were extremely high (0.976 $\pm$ 0.013), and far higher than those among samples (0.3 $\pm$ 0.16).

@@

\subsubsection*{Taxon and Life History Patterns}
We were able to assign a taxon name with confidence to 136 of 146 OTU sequences. The vast majority of sequences (97.6\%) and OTUs (96.9\%) were matched to organisms that have high potential for dispersal at either the gamete, larval, or adult stage, making it impossible to determine whether the source of that DNA was adults with well-documented spatial patterns (e.g. sessile nearshore specialists) or highly mobile early life history stages. Of the 6 OTUs for which dispersal is limited during all life history stages, only 2 occurred in more than two samples, precluding a quantitative comparison of spatial dispersion based on life history characteristics. These were assigned to \textit{Cymatogaster aggregata}, a viviparous nearshore fish with internal fertilization, and \textit{Cupolaconcha meroclista}, a sessile Vermetid gastropod with presumed internal fertilization and short larval dispersal \citep{Strathmann2006, Phillips2010, Calvo2004}. \textit{Cymatogaster aggregata} was distinctly more abundant close to shore, with no sequences occurring in any sample beyond 250 meters (Figure \ref{otu_in_space_select}). \textit{Cupolaconcha meroclista} showed no such distinct spatial trend, occurring in nearly equal abundance at three sites, 75, 500, and 2000 meters from shore. An additional species that was highly abundant in the sequence data, the krill \textit{Thysanoessa raschii}, has pelagic adults, highly seasonal reproduction, and sinking eggs; their distribution was consistent with our expectations based on a tendency of adults to aggregate offshore. Finally, the two most abundant taxa in the dataset were the mussel genus \textit{Mytilus} and the Barnacle order Sessilia; the adults of both taxa are sessile and occur exclusively on hard intertidal substrata but have highly motile larvae.

@@

\subsubsection*{(4) Spatial patterning of eDNA is associated with taxon-specific life history characteristics.}
In contrast to our expectations, other taxa including species with sessile adult stages restricted to benthic hard substrates (e.g. barnacles, mussels) are among the most abundant taxa at sites furthest from shore. However, the larvae and gametes of these taxa are abundant, pelagic, and can be transported long distances by water movement \citep{Strathmann1987}. This indicates that we likely detected DNA of their pelagic phase gametes and larvae. It is always possible that DNA of adults was advected over long distances and detected offshore but in light of our results with krill and surfperch, we view this as unlikely. We interpret our results as evidence that the chaotic spatial distribution of eDNA communities (Figure \ref{pam_in_space}) results from our primers' affinity for many species which at some point exist as microscopic pelagic gametes or larvae. Our results emphasize that expected results based on easily visually observed individuals or detectable with traditional sampling gear such as nets may be very different from results using eDNA. This does caution that eDNA surveys may have different purposes and may not be directly comparable to existing surveys \citep{Shelton2016}.

@@

>>>>

%-----------------------------------NOTES---------------------------------------
% (TODO INSERT DOI FROM FINAL COMMIT BEFORE PUBLICATION)

@@

%-----------------------------------RESULTS-------------------------------------
\section*{Results}
\subsection*{Sequence Data Processing (Bioinformatics)}
Preliminary sequence analysis strongly suggested that the observed variation among environmental samples reflects true variation in the environment, rather than variability due to lab protocols, for the following reasons (note that all value ranges are reported as mean plus and minus one standard deviation). First, all libraries passed the FastQC per-base sequence quality filter, generating a total of 371,576,190 reads passing filter generated in each direction. Second, samples in this study were represented by an adequate number of reads (333,537.9 $\pm$ 112,200.5), with no individual sample receiving fewer than 130,402 reads. Third, there was a very low frequency of cross-contamination from other libraries into those reported here (5e-05$\pm$8e-05; max proportion 0.00034). Fourth, after scaling all samples to the same sequencing depth, OTUs with abundance greater than 178 reads (0.14\% of a sample's reads) experienced no turnover among PCR replicates within a sample. Fifth, sequence abundances among PCR replicates within water samples were remarkably consistent. A single sample had low similarity among PCR replicates (0.659) after removing this outlier, the lowest mean similarity among replicates within a sample was 0.966. Overall similarities among PCR replicates within a sample were extremely high (0.976 $\pm$ 0.013), and far higher than those among samples (0.3 $\pm$ 0.16).
Across PCR replicates, each sample was represented by at least 781425 reads in the raw data and contained between 111 and 443 rarefied OTUs (Supplemental Figure \ref{rarefaction}).

@@

\subsubsection*{Taxon and Life History Patterns}
We were able to assign a taxon name with confidence to 136 of 146 OTU sequences. The vast majority of sequences (97.6\%) and OTUs (96.9\%) were matched to organisms that have high potential for dispersal at either the gamete, larval, or adult stage, making it impossible to determine whether the source of that DNA was adults with well-documented spatial patterns (e.g. sessile nearshore specialists) or highly mobile early life history stages. Of the 6 OTUs for which dispersal is limited during all life history stages, only 2 occurred in more than two samples, precluding a quantitative comparison of spatial dispersion based on life history characteristics. These were assigned to \textit{Cymatogaster aggregata}, a viviparous nearshore fish with internal fertilization, and \textit{Cupolaconcha meroclista}, a sessile Vermetid gastropod with presumed internal fertilization and short larval dispersal \citep{Strathmann2006, Phillips2010, Calvo2004}. \textit{Cymatogaster aggregata} was distinctly more abundant close to shore, with no sequences occurring in any sample beyond 250 meters (Figure \ref{otu_in_space_select}). \textit{Cupolaconcha meroclista} showed no such distinct spatial trend, occurring in nearly equal abundance at three sites, 75, 500, and 2000 meters from shore.
An additional species that was highly abundant in the sequence data, the krill \textit{Thysanoessa raschii}, has pelagic adults, highly seasonal reproduction, and sinking eggs; their distribution was consistent with our expectations based on a tendency of adults to aggregate offshore.
Finally, the two most abundant taxa in the dataset were the mussel genus \textit{Mytilus} and the Barnacle order Sessilia; the adults of both taxa are sessile and occur exclusively on hard intertidal substrata but have highly motile larvae.
Because large-scale dispersal could not be ruled out for the vast majority of taxa, subsetting the community data by taxonomic group had no qualitative effect on the spatial patterning or diversity metrics, and we omit those results here.

@@

\subsubsection*{(4) Spatial patterning of eDNA is associated with taxon-specific life history}
In contrast to our expectations, other taxa including species with sessile adult stages restricted to benthic hard substrates (e.g. barnacles, mussels) are among the most abundant taxa at sites furthest from shore. However, the larvae and gametes of these taxa are abundant, pelagic, and can be transported long distances by water movement \citep{Strathmann1987}. This indicates that we likely detected DNA of their pelagic phase gametes and larvae. It is always possible that DNA of adults was advected over long distances and detected offshore but in light of our results with krill and surfperch, we view this as unlikely. We interpret our results as evidence that the chaotic spatial distribution of eDNA communities (Figure \ref{pam_in_space}) results from our primers' affinity for many species which at some point exist as microscopic pelagic gametes or larvae. Our results emphasize that expected results based on easily visually observed individuals or detectable with traditional sampling gear such as nets may be very different from results using eDNA. This does caution that eDNA surveys may have different purposes and may not be directly comparable to existing surveys \citep{Shelton2016}.

@@

\begin{figure}[H] % [h!] forces the figure to be placed roughly here
  \centering
    \includegraphics[height=0.6\textheight]{rarefaction.pdf}
    \caption{\protect\input{"../../Figures/rarefaction_legend.txt"}}
  \label{rarefaction}
\end{figure}
